A Model to Predict Survival in Patients with End-Stage Liver Disease

### 2022-06-06 analysis/article.Rmd
feat: add first discussion draft

There have been some attempts to use the data extracted from more than 300.000
electronic medical records from two hospitals in the United States to improve
the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9 risk score
add albumin, white blood cell count, age and total cholesterol and length of
stay to the MELD-Na variables. Despite its published prediction improvement the
MELD-Plus scores are not used for transplant allocation yet.

@@

After selecting the final model we train and validated it as described in
@harrell1996.
Briefly we develop our model and predict survival at `r ameldcfg$times` days.
The predictions were grouped into equal sized intervals so that there
are `r ameldcfg$m` patients in each interval.
For each group the mean survival probability and the difference to the
Kaplan-Meier estimate was calculated.
The same procedure was repeated for `r ameldcfg$nboot` bootstrap samples.
Subsequently the differences between predicted survival and Kaplan-Meier
estimates of the bootstrap samples were averaged and added to the original
model difference to obtain a bias-corrected estimated between predicted and
observed survival.

@@

In this retrospective analysis of `r n_analysed` consecutive patients,
who were recruited during the evaluation process for liver transplantation at
the University Hospital of Leipzig we compared
`r n_lrns` different statistical and machine learning algorithms.

...

- higher mortality than predicted, compare @hernaez2020
- adding sodium doesn't add much/any discrimination information
    - as already seen in @kartoun2017
- @kartoun2017 uses > 60 variables mostly clinical stuff, just few laboratory values, none of the clinical stuff remains
- general low sample size thus simpler models/fewer predictors are preferred.
- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features
    - 90 day mortality *post* discharge
    - poorer performance in own validation set (lower and more uniform disease severity, also lower MELD)
- machine learning algorithms worse than statistical learning (lower c-index, higher variance/instability)
    - additive effects (favour statistical models)
    - low sample size, for ML more than 200 events per predictor variable necessary [@ploeg2014]
- national register required

@@

- comparable mortality?
- retrospective analysis
- single center
- unknown cause of death (death with vs of end-stage liver disease)
- bias caused by LTx censorship?

>>>>

There have been some attempts to use the data extracted from more than 300.000
electronic medical records from two hospitals in the United States to improve
the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9 risk score
add albumin, white blood cell count, total cholesterol, age and length of
stay to the MELD-Na variables. Despite its published prediction improvement the
MELD-Plus scores are not used for transplant allocation yet.

@@

After selecting the final model we trained and validated it as described in
@harrell1996.
Briefly we developed our model and predicted survival at `r ameldcfg$times` days.
The predictions were grouped into equal sized intervals so that there
are `r ameldcfg$m` patients in each interval.
For each group the mean survival probability and the difference to the
Kaplan-Meier estimate was calculated.
The same procedure was repeated for `r ameldcfg$nboot` bootstrap samples.
Subsequently the differences between predicted survival and Kaplan-Meier
estimates of the bootstrap samples were averaged and added to the original
model difference to obtain a bias-corrected estimated between predicted and
observed survival.

@@

In this retrospective analysis of `r n_analysed` consecutive patients,
who were recruited during the evaluation process for liver transplantation at
the University Hospital of Leipzig we developed a new model to predict the
90-day survival probability. Our new model outperforms the MELD-Na score and
the recently published MELD-Plus risk score.

**NUMBER_OF_VARIABLES**

- more variables, harder to cheat
- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features

Sodium is not part of our model.
In line with the results presented in @kartoun2017 we could show that
adding sodium to MELD the discrimination performance increases just
insignificantly.

It has shown that women get lower scores in MELD-Na and therefore
are less likely getting an urgent listing for liver transplantation
[@Gaertner2022].
To address sexual differences in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017].
Additionally this transformation yield comparable laboratory measurements
across different clinical laboratories and methods which could improve the
objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample depending scaling prior to model development.
Both should increase the external applicability of our approach.

Additionally to laboratory data we investigated
`r (ncol(raw_data) - length(cn_labs) - 2)`
clinical data but none of them were chosen in our model selection process.
@kartoun2017 used a similar penalizing approach and neither select any clinical
data. Especially in retrospective analysis the clinical data are often not
detailed enough.

In our benchmark of `r n_lrns` different machine learning algorithms we have
seen that the more complex algorithm like boosting and neural networks perform
worse than simpler ones.
This may due additive effects that favour regression models or due to our
relative small sample size.
In machine learning more than 200 events per predictor variable are necessary
to achieve good prediction performance [@ploeg2014].

<!--

- higher mortality than predicted, compare @hernaez2020
- adding sodium doesn't add much/any discrimination information
    - as already seen in @kartoun2017
- @kartoun2017 uses > 60 variables mostly clinical stuff, just few laboratory values, none of the clinical stuff remains
- general low sample size thus simpler models/fewer predictors are preferred.
- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features
    - 90 day mortality *post* discharge
    - poorer performance in own validation set (lower and more uniform disease severity, also lower MELD)
- machine learning algorithms worse than statistical learning (lower c-index, higher variance/instability)
    - additive effects (favour statistical models)
    - low sample size, for ML more than 200 events per predictor variable necessary [@ploeg2014]
- sex-differences and sample-dependencies (by scaling) addressed by *zlog*.
- more variables, harder to cheat
- national register required
-->

@@

Our study has limitations. Principally it is a single center, retrospective
analysis and without an external validation cohort.
The data were collected during the evaluation process for liver transplantation
at the University Hospital of Leipzig.
There was no controlled follow-up process. Instead the follow-up information
were recored at the next clinical admission.
Furthermore we excluded all patients who had already a liver transplantation
which could artificial lower the disease severity of the cohort.

With a median MELD score of `r .tbl1("ScoreMeldUnos")` our
cohort was healthier than in previous studies that reported
a score of 14-15 [@kim2008; @kartoun2017].
However the 90-day mortality rate of `r .tbl1("M90")` was comparable to
the mortality rates seen in earlier studies that ranges from 6 to 30%
[@malinchoc2000; @kim2008; @kartoun2017].
Although our cohort was healthier with a medium mortality rate the classical
MELD score underestimates the observed mortality
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times.
Unfortunately our sample size is very low in the higher MELD categories which
yield a poor but better than MELD based prediction performance of our model
for very ill patients.

In conclusion we developed a new model to predict the 90-day survival of
patients during the evaluation process for liver transplantation.
Our model outperforms the established MELD-Na and the recently published
MELD-Plus risk scores.
For a wider adoption a prospective multi center study for model
recalibration and validation is needed.

<!--
Mortality
[@malinchoc2000] ~ 30 %
[@kim2008] ~ 6-7 %
[@kartoun2017] ~ 16.3 % (90 d, post-discharge-mortality)

random time point for blood sampling

It was not possible to ensure that patients died of end-stage liver disease.

The variables we found and the model itself have to be validated and
recalibrated in a multi center, prospective trial.

Few patients with high MELD scores, poor calibration for high mortality

- comparable mortality?
- retrospective analysis
- single center
- unknown cause of death (death with vs of end-stage liver disease)
- bias caused by LTx censorship?
-->
